Fast Sorting on Modern Computers

### 2005-05-28 paper/docs/background.tex
after spell check
spell checked updateed intro read through method removed extras, which was part of the demo added labels

\section{Sorting}

@@

Sorts are frequently classified as \n{in-place} and \n{out-of-place}. An
out-of-place sort typically uses an extra array of the same size as the array to
be sorted. This may require the results to be copied back at the end. An
in-place sort does not, though it may use a large stack, especially in recursive
algorithms.

A \n{stable} sort is one in which records which have the same key stay in the
same order across the sort. All the elementary sorts are stable, as are
mergesort and radixsort. Quicksort, heapsort and shellsort are not stable.

@@

The ADT interface in this project is taken from \cite{Sedgewick02}. An \cc{Item}
is a record, and a function is defined to extract its key.  Comparisons are done
via the \cc{less} function, and swapping is done using the \cc{exch} function.
These are replaced with their actual functions by the C preprocessor. The code
used for ADTs is in Figure \vref{Definitions used by the ADT}. Items in this
project were \cc{unsigned integer}s, and were compared using a simple
\n{less-than} function\footnote{This abstraction broke down in several cases,
such as where a \n{greater-than} function was required, such as in quicksort}.
In this case, a record and key are equivalent, and the word \n{key} is used to
refer to items to be sorted.

@@

Sorts which are not comparison based also exist. These sorts are called counting
based sorts, as they count the number of keys to be put in each segment of an
array before putting them in place. The most important of these is radixsort,
discussed in Chapter \ref{Radixsort}.

@@

\section{Caches}
Processors run many times faster than main memory. As a result, a processor
which must constantly wait for memory is not used to its full potential. To
overcome this problem, a \n{cache hierarchy} is used. A cache contains a small
subset of main memory in a smaller memory, which is much faster to access than
main memory. Modern computers typically contain many levels of cache, each with
increasing size and \n{latency}\footnote{Latency is the time between a request
being made, and its successful execution.}. For example, it may take twice as
long to access the level 1 cache as it does to access a register. It may take
ten times as long as that to access the level 2 cache, and fifty times as long
to access main memory.

@@

\subsubsection{Two-Level Adaptive}
A two-level adaptive predictor uses branch history to predict future branches. A
global branch history register keeps track of the last few branches, and uses
this to index a table of 2-bit predictors, which predict the result of the
branch. Frequently, the program counter is also included, either concatenated or
exclusive-ored with the branch history to form the table index.

@@

An algorithm which is $O(N)$ is said to complete in linear time. An algorithm
which is $O(N^2)$ is said to complete in quadratic time.  $O(NlogN)$ algorithms
may take several orders of magnitude more time than $O(N)$ algorithms, and
typically use the \n{divide-and-conquer} principle, or use a binary tree to
reduce complexity. Quicksort, mergesort and heapsort are all of this type,
whereas elementary sorts, such as insertion sort, selection sort, bubblesort and
shakersort are $O(N^2)$ algorithms. Radixsort is $O(N)$, but can be more
expensive than quicksort, due to the cost of each step.  These are discussed in
later chapters.

>>>>

\label{background}
\section{Sorting}

@@

Sorts are frequently classified as \n{in-place} and \n{out-of-place}. An
out-of-place sort typically uses an extra array of the same size as the array to
be sorted. This may require the results to be copied back at the end. An
in-place sort does not require this extra array, though it may use a large
stack, especially in recursive algorithms.

A \n{stable} sort is one in which records which have the same key stay in the
same relative order during the sort. All the elementary sorts are stable, as are
mergesort and radixsort. Quicksort, heapsort and shellsort are not stable.

@@

The ADT interface in this project is taken from \cite{Sedgewick02}. An \cc{Item}
is a record, and a function is defined to extract its key.  Comparisons are done
via the \cc{less} function, and swapping is done using the \cc{exch} function.
These are replaced with their actual functions by the C preprocessor. The code
used for ADTs is in Figure \vref{Definitions used by the ADT}. The tests
presented in this report use \cc{unsigned integer}s as Items, which were
compared using a simple \n{less-than} function\footnote{This abstraction broke
down in several cases, such as where a \n{greater-than} function was required,
i.e. in quicksort.}.  In this case, a record and key are equivalent, and the
word \n{key} is used to refer to items to be sorted.

@@

Sorts which are not comparison based also exist. These sorts are called counting
based sorts, as they count the number of keys to be put in each segment of an
array before putting them in place. The most important of these is radixsort,
discussed in Chapter \ref{radix}.

@@

\section{Caches}
Processors run many times faster than main memory. As a result, a processor
which must constantly wait for memory is not used to its full potential. To
overcome this problem, a \n{cache hierarchy} is used. A cache contains a small
subset of main memory in a smaller memory, which is much faster to access than
main memory. Modern computers typically contain many levels of cache, each with
increasing size and \n{latency}\footnote{Latency is the time between a request
being made and its successful execution.}. For example, it may take twice as
long to access the level 1 cache as it does to access a register. It may take
ten times as long as that to access the level 2 cache, and fifty times as long
to access main memory.

@@

\subsubsection{Two-Level Adaptive}
A two-level adaptive predictor uses branch history to predict future branches. A
global branch history register keeps track of the last few branches, and uses
this to index a table of 2-bit predictors, which predict the result of the
branch. Frequently, the program counter is also included, either concatenated or
exclusive-ORed with the branch history to form the table index.

@@

An algorithm which is $O(N)$ is said to complete in linear time. An algorithm
which is $O(N^2)$ is said to complete in quadratic time.  $O(NlogN)$ algorithms
may take several orders of magnitude more time than $O(N)$ algorithms, and
typically use the \n{divide-and-conquer} principle, or use a binary tree to
reduce complexity. Quicksort, mergesort and heapsort are all of this type,
whereas elementary sorts, such as insertion sort, selection sort, bubblesort and
shakersort are $O(N^2)$ algorithms. Radixsort is $O(N)$, but can be more
expensive than quicksort, due to the cost of each step.  These are discussed in
later chapters.
%TODO what's shellsort in Big O notation
TODO what's shellsort in Big O notation

### 2005-05-28 paper/docs/conclusions.tex
spell checked updateed intro read through method removed extras, which was part of the demo added labels

This section presents a summary of the results from previous chapters, lists
important results and discoveries, and suggests contributions made to computer
science by this project.

>>>>

\label{conclusions}
This section presents a summary of the results from previous chapters, lists
important results and discoveries, and suggests contributions made to computer
science by this project.

### 2005-05-28 paper/docs/heap.tex
spell checked updateed intro read through method removed extras, which was part of the demo added labels

%\section{Description}

@@

\subsubsection{Out-of-place sorting}
Up until this point, the heap property used has been that a parent will be
larger than its child. If this property is change so that the parent is always
smaller than its child, then the result is an out-of-place sort. The smallest
key is removed from the heap, and placed in its final place in the array. The
heap is then re-heapified. LaMarca refers to the steps involved as
\n{Remove-min}.

\section{Base heapsort}
LaMarca recommends to use of the Repeated-Adds algorithm for reducing
cache misses. As such, his base heapsort uses Floyd's method, ostensibly at
the behest of major textbooks, each of whom claim Floyd's method gives better
results. This is certainly backed up by Sedgewick, who recommends this method,
but in initial tests, the out-of-place Repeated-Adds technique had a lower
instruction count. Regardless, the base algorithm used here is Floyd's in-place
sort. Code for a base heapsort is in Figure \ref{Simple Base Heapsort}.

>>>>

\label{heap}
%\section{Description}

@@

\subsubsection{Out-of-place Sorting}
Up until this point, the heap property used has been that a parent will be
larger than its child. If this property is change so that the parent is always
smaller than its child, then the result is an out-of-place sort. The smallest
key is removed from the heap, and placed in its final place in the array. The
heap is then re-heapified. LaMarca refers to the steps involved as
\n{Remove-min}.

\section{Base Heapsort}
LaMarca recommends to use of the Repeated-Adds algorithm for reducing
cache misses. As such, his base heapsort uses Floyd's method, ostensibly at
the behest of major textbooks, each of whom claim Floyd's method gives better
results. This is certainly backed up by Sedgewick, who recommends this method,
but in initial tests, the out-of-place Repeated-Adds technique had a lower
instruction count. Regardless, the base algorithm used here is Floyd's in-place
sort. Code for a base heapsort is in Figure \ref{Simple Base Heapsort}.

### 2005-05-28 paper/docs/intro.tex
spell checked updateed intro read through method removed extras, which was part of the demo added labels

Most major sorting algorithms in computer science date to the 50s and 60s.
Radixsort was written in 1954, quicksort in 1962, and mergesort has been traced
back to the 1930s. Modern processors meanwhile have features which were not
thought of when these sorts were written: branch predictors, for example, were
not created until the 1970s, and advanced branch prediction techniques were
not used in consumer processors until very recently. It is hardly surprising,
then, that sorting algorithms rarely take these architectural features into
account.

Papers considering sorting performance typically analyse instruction count.
LaMarca and Ladner \cite{LaMarca96} break this trend by discussing the cache
implications of several algorithms, including sorts. This project attempts to
investigate these claims and to test and analyse the high performance sorts
based on them.

LaMarca devotes a chapter of his thesis to sorting. A chapter is spent
discussing implicit heaps, which are later used in heapsort. In
\cite{LaMarca99}, radixsort was added for comparison. Radixsort is added here
for the same purpose. For comparison, elementary sorts - insertion sort,
selection sort, bubblesort and shakersort - are also included here.

In addition to repeating LaMarca's work, this project also discusses hardware
branch prediction techniques and their effects on sorting. When algorithms are
altered to improve their performance, it is important to observe if any change
in the rate of branch prediction misses occurs.

@@

Several important results are discovered during the course of the project. The
branch prediction rates of insertion sort and selection sort, for example, and
the difference in performance between binary searches and sequential searches due
to branch misses. Several steps to remove instructions from standard algorithms
are devised, such as double tiling mergesort, and removing a sentinel and bounds
check from heapsort.

Chapter 2 provides background information on caches and branch predictors. It
also discusses the theory of sorting, and conventions used in the algorithms in
this project.

Chapter 3 discusses the method and framework used in this project, including the
tools used and the reasons for using them.

Chapter 4 discusses the elementary sorts and their implementations. Results are
included for later comparison.

Chapters 5 through 7 discuss base algorithms and cache-conscious improvements
for heapsort, mergesort and quicksort. Results are presented and explained, and
branch prediction implications of the changes are discussed.

Chapter 8 presents a radixsort implementation and discusses its performance in
the areas of instruction count, cache-consciousness and branch prediction.

Chapter 9 presents conclusions and contributions developed over the course of
this project.

>>>>

\label{intro}
Most major sorting algorithms in computer science date to the 50s and 60s.
Radixsort was written in 1954, quicksort in 1962, and mergesort has been traced
back to the 1930s. Modern processors meanwhile have features which were not
thought of when these sorts were written: branch predictors, for example, were
not created until the 1970s, and advanced branch prediction techniques were
not used in consumer processors until very recently. It is hardly surprising,
then, that sorting algorithms rarely take these architectural features into
account.

Papers considering sorting performance typically analyse instruction count.
LaMarca and Ladner \cite{LaMarca96} break this trend by discussing the cache
implications of several algorithms, including sorts. This report 
investigates these claims and tests and analyses high performance sorts
based upon thier work.

LaMarca devotes a chapter of his thesis to sorting. Another chapter is spent
discussing implicit heaps, which are later used in heapsort. In
\cite{LaMarca99}, radixsort was added for comparison. Radixsort is added here
for the same purpose. For comparison, elementary sorts - insertion sort,
selection sort, bubblesort and shakersort - are also included here.

In addition to repeating LaMarca's work, this report also discusses hardware
branch prediction techniques and their effects on sorting. When algorithms are
altered to improve their performance, it is important to observe if any change
in the rate of branch prediction misses occurs.

@@

Several important discovered results are reported here. The
branch prediction rates of insertion sort and selection sort, for example, and
the difference in performance between binary searches and sequential searches due
to branch misses. Several steps to remove instructions from standard algorithms
are devised, such as double tiling mergesort, removing a sentinel and bounds
check from heapsort.

Chapter \ref{background} provides background information on caches and branch predictors. It
also discusses the theory of sorting, and conventions used in the algorithms in
this project.

Chapter \ref{method}  discusses the method and framework used in this project, including the
tools used and the reasons for using them.

Chapter \ref{ordernsquared}  discusses the elementary sorts and their implementations. Results are
included for later comparison.

Chapters \ref{heap} through \ref{quick} discuss base algorithms and cache-conscious improvements
for heapsort, mergesort and quicksort. Results are presented and explained, and
branch prediction implications of the changes are discussed.

Chapter \ref{radix} presents a radixsort implementation and discusses its performance in
the areas of instruction count, cache-consciousness and branch prediction.

%TODO shell
TODO Chapter \ref{shell}

Chapter \ref{conclusions} presents conclusions, and lists contributions
described by this report.

### 2005-05-28 paper/docs/merge.tex
spell checked updateed intro read through method removed extras, which was part of the demo added labels

%\section{Description}
\label{Mergesort}
Merging is a technique which takes two sorted lists, and combines them into a
single sorted list. Starting at the smallest key on both lists, it writes the
smaller key to a third list. It then compares the smallest key on one list with
the next smallest on the other, and writes the smaller to the auxiliary list.
The merging continues until both lists are combined into a single sorted list.

@@

\subsection{Base Mergesort}
Three optimisations are described by LaMarca for the base mergesort. Firstly, a
bitonic sort, as described in \cite{Sedgewick02}, removes instructions from the
inner loop. Every second array is created in reverse, so that the larger keys  
are next to each other in the middle. The smallest keys are at the outside,
meaning that the lists are iterated through from the outside in. As a result, it
is not possible to go outside the array, and it is not necessary to consider the
lengths of the arrays, apart from when setting the initial array indices. When
the first array is exhausted, it will point to the last key on the other array,
which will be larger than all other keys in that array. When both are
exhausted, the last key of both arrays will be equal. A check is put in, in this
case, and merging ends if both indices are the same. Figure \vref{Visual Sort}
shows a bitonic merge.

>>>>

%\section{Description}
\label{merge}
Merging is a technique which takes two sorted lists, and combines them into a
single sorted list. Starting at the smallest key on both lists, it writes the
smaller key to a third list. It then compares the smallest key on one list with
the next smallest on the other, and writes the smaller to the auxiliary list.
The merging continues until both lists are combined into a single sorted list.

@@

\subsection{Base Mergesort}
\label{base mergesort}
Three optimisations are described by LaMarca for the base mergesort. Firstly, a
bitonic sort, as described in \cite{Sedgewick02}, removes instructions from the
inner loop. Every second array is created in reverse, so that the larger keys  
are next to each other in the middle. The smallest keys are at the outside,
meaning that the lists are iterated through from the outside in. As a result, it
is not possible to go outside the array, and it is not necessary to consider the
lengths of the arrays, apart from when setting the initial array indices. When
the first array is exhausted, it will point to the last key on the other array,
which will be larger than all other keys in that array. When both are
exhausted, the last key of both arrays will be equal. A check is put in, in this
case, and merging ends if both indices are the same. Figure \vref{Visual Sort}
shows a bitonic merge.

### 2005-05-28 paper/docs/method.tex
spell checked updateed intro read through method removed extras, which was part of the demo added labels

This section discusses the method and framework used to complete this project.
From the point of view of a Final Year Project, this is a description of what I
did all year, what I spent my time on, what went right and wrong during the
project, and what I learned. The first section discusses the steps involved in
creating the project, programming the sorts and getting results. The second
section discusses tools that I used or created in the course of the project.

\section{Method}

The following is the initial project description:
\begin{quotation}
Sorting is one of the most important and studied problems in computer science.
Many good algorithms exist which offer various trade-offs in efficiency,
simplicity, and memory use.  One weakness of classical sorting algorithms is
that they are not designed for modern computers.

@@

Once the algorithm was written, a testing framework was used. This tested for
edge-cases in the algorithms by testing sorts sized from $8$ to $255$ keys, with
different random data sets. Testing was also performed on a very large data set.
This test was timed and the results were considered. The results of this test
were not accurate: the program took up to a minute to run, during which time the
computer was being used by multiple users and processes, leading to context
switches,sharing of resources and scheduling. Despite these inaccuracies, it was
possible to use the results relative to each other, and to determine whether,
for example, memory-tuned quicksort was running faster than base quicksort.

@@

This shows two stages of a bitonic mergesort. The top of the image shows the
source array being merged into the target array below. The source array shows
three large segments of the array and four smaller segments. The first two large
segments have been merged into one segment in the target array. The third of
these has been merged into the corresponding target array segment. The first of
the smaller segments has begun being merged with the other small segments. The
rest of the target array is uninitialised.  Below that is a small amount of
debugging information. $i == j$ is a key point in the program and is printed as
debugging information. In the two arrays below this, the merge continues.

@@

To show an example of the creation of a sort, mergesort will now be discussed.
Descriptions of some of the terms here is saved for discussion in Chapter
\ref{Mergesort}.

@@

A tiled mergesort was then written, based on LaMarca's guidance. Firstly, this
was aligned in memory so that the number misses due to conflicts would be
reduced.  Then it fully sorted segments of the array, each half the size of the
cache, so that temporal locality would be exploited. Finally, all these segments
were merged together in the remaining merge steps.

@@

Once these sorts were written it was attempted to improve them slightly.
Instead of a tiled mergesort, a double tiled mergesort was written.  This
attempted to completely sort in the level 1 cache, then to completely sort in
the level 2 cache. Double multi-mergesort was a combination of this and of
multi-mergesort. The rewrite was based on the base mergesort, and was written
far more cleanly, being more readable and efficient. These were also found to
reduce the instruction count, as well as the level 2 cache miss count. These
results are discussed in more detail in \ref{Mergesort results}.

@@

\section{Future work}
SimpleScalar ran very slowly, with each simulation run taking at least $2
\frac{1}{2}$ hours. Several steps could be taken to lower this time. Firstly,
compiling SimpleScalar with the \cc{-03} flag, which performs a significant
number of optimisations on the program. This was not done because of the
difficulty of compiling the program in the first place.

\cc{lib-cheetah} is a SimpleScalar library used to simulate multiple
caches in the same simulation. Over half the simulation time was taken by four
cache simulations, and running them concurrently might have significantly
decreased the running time. This was not a significant worry during this project,
but decreasing the running time would be necessary if the either the number of
simulations or the number of sorts was to change dramatically.

An improvement that would have yielded interesting results would be to do simulations
which only consisted of flow control, with no comparisons. This would allow a
division of flow control misses and comparative misses. However, this would not
always be possible; sorts like quicksort and heapsort mix use their data to
control the flow of the program. In addition, bubblesort, shakersort and
selection sort have (analytically) predictable flow control, while radixsort has
no misses at all. Mergesort may benefit from this, however, especially versions
with difficult flow control diagrams, such as \n{algorithm N} and \n{algorithm
S} (see Section \ref{Algorithm N}).

>>>>

\label{method}
%TODO rewrite this
This section discusses the method and framework used to complete this project.
From the point of view of a Final Year Project, this is a description of what I
did all year, what I spent my time on, what went right and wrong during the
project, and what I learned. The first section discusses the steps involved in
creating the project, programming the sorts and getting results. The second
section discusses tools that I used or created in the course of the project.

\section{Method}
i
%TODO remove this
The following is the initial project description:
\begin{quotation}
Sorting is one of the most important and studied problems in computer science.
Many good algorithms exist which offer various trade-offs in efficiency,
simplicity, and memory use.  One weakness of classical sorting algorithms is
that they are not designed for modern computers.

@@

Once the algorithm was written, a testing framework was used. This tested for
edge-cases in the algorithms by testing sorts sized from $8$ to $255$ keys, with
different random data sets. Testing was also performed on a very large data set.
This test was timed and the results were considered. The results of this test
were not accurate: the program took up to a minute to run, during which time the
computer was being used by multiple users and processes, leading to context
switches, sharing of resources and scheduling. Despite these inaccuracies, it was
possible to use the results relative to each other, and to determine whether,
for example, memory-tuned quicksort was running faster than base quicksort.

@@

This shows two stages of a bitonic mergesort which is described in Section \ref{base
mergesort}. The top of the image shows the source array being merged into the
target array below. The source array shows three large segments of the array and
four smaller segments. The first two large segments have been merged into one
segment in the target array. The third of these has been merged into the
corresponding target array segment. The first of the smaller segments has begun
being merged with the other small segments. The rest of the target array is
uninitialised.  Below that is a small amount of debugging information. $i == j$
is a key point in the program and is printed as debugging information. In the
two arrays below this, the merge continues.

@@

To show an example of the creation of a sort, mergesort will now be discussed.
Descriptions of some of the terms here is saved for discussion in Chapter
\ref{merge}.

@@

A tiled mergesort was then written, based on LaMarca's guidance. Firstly, this
was aligned in memory so that the number of misses due to conflicts would be
reduced.  Then it fully sorted segments of the array, each half the size of the
cache, so that temporal locality would be exploited. Finally, all these segments
were merged together in the remaining merge steps.

@@

Once these sorts were written it was attempted to improve them slightly.
Instead of a tiled mergesort, a double tiled mergesort was written.  This
attempted to completely sort in the level 1 cache, then to completely sort in
the level 2 cache. Double multi-mergesort was a combination of this and of
multi-mergesort. The rewrite was based on the base mergesort, and was written
far more cleanly, being more readable and efficient. These were also found to
reduce the instruction count, as well as the level 2 cache miss count. These
results are discussed in more detail in Section \ref{Mergesort results}.

@@

\section{Future work}
An improvement that would have yielded interesting results would be to do simulations
which only consisted of flow control, with no comparisons. This would allow a
division of flow control misses and comparative misses. However, this would not
always be possible; sorts like quicksort and heapsort use their data to
control the flow of the program. In addition, bubblesort, shakersort and
selection sort have (analytically) predictable flow control, while radixsort has
no comparison misses at all. Mergesort may benefit from this, however, especially versions
with difficult flow control diagrams, such as \n{algorithm N} and \n{algorithm
S} (see Section \ref{Algorithm N}).
